---
title: "day6-wrangling"
author: "Meagan Brown"
date: '2022-08-15'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(here)
```

## Read in the world bank data files

```{r}
wb_indicators <- read_csv(here("data", "wb_indicators.csv"), 
                          na = c("..", "")) 
```

## Tidy up my data

```{r}

wb_indicators_long <- wb_indicators %>% 
  pivot_longer(cols = '2001 [YR2001]':'2020 [YR2020]',
               names_to = "year",
               values_to = "indicator_value")

```

### Separate to clean up year column

```{r}
wb_clean <- wb_indicators_long %>% 
  separate(col = year,
           into = c("year",
                    "year_chr"),
           sep = " ") %>% 
  select(-year_chr, 
         -'Country Code', 
         -'Series Code') %>% 
  mutate(year = as.numeric(year)) %>% 
  drop_na('Series Name')
```

## Pivoting Wider

```{r}
wb_indicators_wide <- wb_clean %>% 
  pivot_wider(names_from = 'Series Name',
              values_from = 'indicator_value') 
```

## Renaming

```{r}
names(wb_indicators_wide) <- c("country",
                     "year",
                     "acces_clean_fuels_pp",
                     "access_electricity_pp",
                     "co2_emissions_kt",
                     "fossil_fuel_cons_pct",
                     "water_stress")

wb_subset <- wb_indicators_wide %>% 
  filter(country %in% c("Algeria",
                        "Barbados",
                        "Bulgaria",
                        "Chile"))

ggplot(data = wb_subset,
       aes(x = year,
           y = co2_emissions_kt,
           group = country)) +
  geom_line(aes(color = country)) +
  facet_wrap(~country)
```

## filtering rows

Example 1: Filter to only keep exact matches

```{r}
ex_1 <- starwars %>% 
  filter(height > 180)
```

```{r}
ex_2 <- starwars %>% 
  filter(eye_color == "blue")
```

```{r}
ex_3 <- starwars %>% 
  filter(homeworld == "Naboo")
```

```{r}
ex_4 <- starwars %>% 
  filter(height > 180 & homeworld == "Tatooine")
```

```{r}
ex_5 <- starwars %>% 
  filter(hair_color == "brown" & species == "Human")

```

```{r}
ex_6 <- starwars %>% 
  filter(height > 180 | eye_color == "yellow")
```

```{r}
ex_7 <- starwars %>% 
  filter(species == "Droid" | homeworld == "Endor")
```

```{r}
ex_8 <- starwars %>% 
  filter(eye_color %in% c("blue",
                          "red",
                          "yellow"))

ex_10 <- starwars %>% 
  filter(homeworld %in% c("Endor", 
                          "Tatooine", 
                          "Naboo"),
         height < 150)

ex_12 <- starwars %>% 
  filter(species != "human")

ex_13 <- starwars %>% 
  filter(!species %in% c("Human",
                         "Ewok",
                         "Wookie"))
```

## 'select()': select or exclude columns

```{r}

ex_14 <- wb_indicators_wide %>% 
  select(year:fossil_fuel_cons_pct)

ex_16 <- wb_indicators_wide %>% 
  select(year:fossil_fuel_cons_pct, -access_electricity_pp)

```

```{r}
ex_17 <- wb_indicators_wide %>% 
  filter(country %in% c("Bolivia",
                        "Chile",
                        "Mexico")) %>% 
  select(country:fossil_fuel_cons_pct)
```

```{r}
ex_18 <- wb_indicators_wide %>% 
  select(year_new = year,
         emissions = co2_emissions_kt,
         country_name = country) 

ex_19 <- wb_indicators_wide %>% 
  select(year, 
         country, 
         water_stress) %>% 
  filter(country %in% c("Australia", 
                        "Aruba", 
                        "Lebanon")) %>% 
  drop_na()

```

### `rename`

```{r}
ex_20 <- wb_indicators_wide %>% 
  rename(ffcp = fossil_fuel_cons_pct,
         ws = water_stress)
```

```{r}
ex_21 <- wb_indicators_wide %>% 
  rename(acfpp = access_electricity_pp)
```

### `mutate()`: update an existing column, or add a new one

If you assign a NEW column name, it will add a new column to the end of your data frame. IF you assign an existing column name it will overwrite that column.

```{r}
#converting emissions from kt to t
ex_22 <- wb_indicators_wide %>% 
  mutate(co2_emissions_t = co2_emissions_kt * 1000) %>% 
  relocate(co2_emissions_t, .after = co2_emissions_kt)

#creating a new column named `yay` with values `YAY`
ex_23 <- wb_indicators_wide %>% 
  mutate(yay = "YAY")

#converting to charcater
ex_24 <- wb_indicators_wide %>% 
  mutate(year = as.character(year))

#converting to numeric
ex_25 <- ex_24 %>% 
  mutate(year = as.numeric(year))

#relocating water_stress column to be after year column, can also use .before
ex_26 <- wb_indicators_wide %>% 
  relocate(water_stress, 
           .after = year)
```

### `group_by` and `summarize()`

`group_by`: Creates groups under the hood based on variables you specify

`summarize()`: apply yout funvtions to EACH GROUP, and return them in a nice summary table

```{r}
ex_27 <- starwars %>% 
  filter(homeworld %in% c("Naboo", "Tatooine")) %>% 
  group_by(species, homeworld) %>% 
  summarize(mean_height = mean(height, na.rm = TRUE),
            mean_mass = mean(mass, na.rm = TRUE))
```

## Case table (each observation has its own row)
```{r}
df <- tribble(
  ~species, ~length, ~number,
  "lobster", 12, 4,
  "lobster", 14, 6,
  "lobster", 15, 2
)

df_case <- df %>% 
  uncount(number)

# Count function counts observations
# Combination of group_by summarize n()
starwars %>% 
  count(homeworld, species) 

starwars %>% 
  group_by(homeworld, species) %>% 
  summarize(size = n(),
            mean_height = mean(height, na.rm = TRUE))

```

```{r}
practice <- wb_indicators_wide %>% 
  filter(country %in% c("Nicaragua",
                        "Ecuador",
                        "Peru",
                        "Chile"),
                       year > 2005) %>% 
  select(country, year, co2_emissions_kt) %>% 
  mutate(co2_emissions_mt = co2_emissions_kt / 1000) %>% 
  group_by(country) %>% 
  summarize(co2_emissions_mt = sum(co2_emissions_mt, na.rm = TRUE))
  
```

